---
title: "Bayse'sche Item-Reponse-Modellierung mit R"
author: "Simon Schäfer"
date: "15 8 2021"
output: ioslides_presentation
widescreen: true
incremental: true
logo: Hu_Logo_small.png
css: ioslidesstyle.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.12.2/jquery.min.js"></script>

<script>
    $(document).ready(function() {
      $('slide:not(.title-slide, .backdrop, .segue)').append('<footer label=\"Simon Schäfer, HU-Berlin\"></footer>');    
    })
</script>

## Überblick

<div style="float: left; width: 40%;">
### Was 

1. Hallo
1. Nein
</div>

<div style="float: right; width: 60%;">
### Was nicht

bla
</div>

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

>- Bullet 1
>- Bullet 2
>- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

